A formal treatment of bidirectional typing

### 2023-10-15 tex/ESOP24/S1-intro.tex
[ tex ] Revised S1 & S4

Type inference is an important mechanism for the transition to well typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\varcitet{Dunfield2021}{ in their survey paper} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
While it is straightforward to derive a bidirectional type synthesis algorithm, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.

\subsubsection{Proof-relevance in type synthesis}
\label{sec:PLFA}

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability implies algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (and can be strengthened to the original form).
Hence, proving the statement is the same as constructing a verified proof-relevant type inferencer, which returns not only an answer but also a proof justifying that answer.

@@

In a bit more detail:
The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser decides the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability described by \citet[Section~3.2]{Dunfield2021}.

@@

\subsubsection{Mode-correctness and general definitions of languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
It seems that the implications of mode-correctness have only been addressed informally as exhibited in the survey paper~\cite{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
%But, what is a language anyway?

To make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The latter theorem effectively amounts to a generator of proof-relevant bidirectional type synthesiser for all syntax-directed bidirectional simple type systems that are mode-correct.

To make our exposition accessible we have chosen to work with simply typed languages so that bidirectional type synthesis is established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
Nevertheless, we briefly discuss a possible extension which includes polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than annotatability~\cite{Dunfield2021}), and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.

>>>>

Type inference is an important mechanism for the transition to well typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\varcitet{Dunfield2021}{ summarised in their survey paper} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis.
%this situation is in stark contrast to parsing, for which there are general theories and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
%While it is usually straightforward to derive a bidirectional type synthesis algorithm,
\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should also start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
%Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
In this paper, we develop such a theory of bidirectional typing with the proof assistant \Agda.
%and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
%In particular, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

\subsubsection{Proof-relevant type synthesis}
\label{sec:PLFA}

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is also algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.

@@

To give a bit more detail:
The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than \varcitet[Section~3.2]{Dunfield2021}{'s annotatability}.

@@

\subsubsection{Mode-correctness and general definitions of languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
It seems that the implications of mode-correctness have only been addressed informally by \citet{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
%But, what is a language anyway?

To make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator of proof-relevant type-synthesisers.
\end{enumerate}
Our theory is fully formally developed in \Agda but translated to the mathematical vernacular for presentation in this paper.
The formal theory doubles as a verified implementation, with no gap in between.

### 2023-10-15 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Update Future work
[ tex ] Revised S1 & S4

Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
We can also remove typing and retain mode information, arriving at a mode derivation instead.
Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
%Formulating the notion of mode derivation helps to complete this clean picture.
Technically, all these can be summarised in one theorem.

@@

The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{To simplify the presentation, we use ordinary mode decoration elsewhere in this paper.}

@@

\begin{theorem}[Generalised mode preprocessing]\label{thm:generalised-mode-decoration}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{theorem}

@@

\begin{proof}[of \cref{thm:generalised-mode-decoration}]
By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.

\begin{itemize}
\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
\item $\Rule{Anno}$:
Let $t = (t' \bbcolon A)$.
By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some $g$~and~$s$, to which we apply $\SynRule{Anno}$.
\item $\Rule{Op}$:
Let $t = \tmOpts$.
By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
\end{itemize}
\end{proof}

@@

\begin{corollary}[Mode preprocessing]\label{thm:mode-decoration}
  It is decidable whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
\end{corollary}

@@

\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}
              {\isTerm{x} \sqsupseteq \isTerm{x}}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
              {\tmOpts \sqsupseteq \tmOpus}
  \end{mathpar}

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved with generalised mode decoration.

\begin{proof}[of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\end{itemize}
\end{proof}

>>>>

Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
We can also remove typing and retain mode information, arriving at a mode derivation instead.
Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
%Formulating the notion of mode derivation helps to complete this clean picture.
Technically, all these can be summarised in one theorem (proved by routine induction).

@@

The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{For the sake of simplicity, we use ordinary mode decoration elsewhere in this paper.}

@@

\begin{theorem}[Generalised mode decoration]\label{thm:generalised-mode-decoration}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{theorem}

@@

%\begin{proof}[of \cref{thm:generalised-mode-decoration}]
%By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.
%
%\begin{itemize}
%\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
%\item $\Rule{Anno}$:
%Let $t = (t' \bbcolon A)$.
%By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some $g$~and~$s$, to which we apply $\SynRule{Anno}$.
%\item $\Rule{Op}$:
%Let $t = \tmOpts$.
%By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
%\end{itemize}
%\end{proof}

@@

\begin{corollary}[Mode decoration]\label{thm:mode-decoration}
  It is decidable whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
\end{corollary}

@@

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.

\begin{figure}[h]
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}
              {\isTerm{x} \sqsupseteq \isTerm{x}}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
              {\tmOpts \sqsupseteq \tmOpus}
  \end{mathpar}

### 2023-10-15 tex/ESOP24/S8-future.tex
[ tex ] Revised S1 & S4

However, explicit type application is impractical but implicit application results in the \emph{instantiation problem} which amounts to replacing type equality checks $A = B$ with checking subtyping $A <: B$ and guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$ in the presence of universal quantification.
Developing a general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.

>>>>

However, explicit type application is impractical but implicit application results in the \emph{instantiation problem} which amounts to replacing type equality checks $A = B$ with checking subtyping $A <: B$ and guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$ in the presence of universal quantification.
A general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.
